Applying Step Selection Functions to Data with Data Gaps: Simulation and Case Study on African wild dogs

### 2022-02-22 Chapter_1/04_Manuscript/Manuscript.tex
updates

\newpage
\begin{abstract}
%------------------------------------------------------------------------------
%  LONG VERSION
%------------------------------------------------------------------------------
Dispersal of individuals contributes to long-term population persistence, yet
requires a sufficient degree of landscape connectivity. To date, connectivity
has mainly been investigated using least-cost analysis and circuit theory, two
methods that make assumptions that are hardly applicable to dispersal.
Least-cost analysis assumes that animals move towards a known endpoint and are
knowledgeable about the most favorable route. Circuit theory presumes a complete
random walk and fails to incorporate directional persistence. While these
assumptions can be relaxed by explicitly simulating dispersal trajectories
across the landscape, a unified approach for such simulations is lacking.

Here, we present a simple three-step approach to simulate dispersal and to
assess connectivity using empirical GPS movement data and a set of habitat
covariates. In step one, we use integrated step-selection functions to fit a
mechanistic movement model describing habitat and movement preferences of
dispersing individuals. In step two, we apply the parameterized model to
simulate dispersal across the study area. In step three, we derive a set of
three complementary connectivity maps; a heatmap highlighting frequently
traversed areas, a betweenness map pinpointing dispersal corridors, and a map of
inter-patch connectivity indicating the presence and intensity of functional
links between habitat patches. As a case study, we apply the three-step
approach to GPS data collected on dispersing African wild dogs (\textit{Lycaon
pictus}) inhabiting the Kavango-Zambezi Transfrontier Conservation Area
(KAZA-TFCA) in southern Africa.

Using step-selection functions we successfully parametrized a detailed dispersal
model that described dispersing individuals habitat and movement preferences, as
well as potential interactions among the two. The model substantially
outperformed a model that omitted such interactions and enabled us to simulate
80,000 dispersal trajectories across the study area. By explicitly simulating
individuals, this approach not only requires fewer unrealistic assumptions about
dispersal, but also permits the calculation of a rich suite of connectivity
metrics that together provide a comprehensive view on landscape connectivity.
The three derived connectivity maps revealed several dispersal hotspots and
corridors across the extent of our study area, yet each map highlighted a
different aspect of landscape connectivity, thus emphasizing their complementary
nature. Overall, our case study demonstrates that a simulation-based approach
offers a simple yet powerful alternative to traditional connectivity modeling
techniques. The approach is therefore useful for a variety of applications in
ecological, evolutionary, and conservation research.

@@

% Importance of Dispersal & Connectivity
\subsection{Importance of Connectivity \& Connectivity Models}
Dispersal of individuals is a vital process that allows species to maintain
genetic diversity \citep{Perrin.1999, Perrin.2000, Frankham.2002, Leigh.2012,
Baguette.2013}, rescue non-viable populations \citep{Brown.1977}, and to
colonize unoccupied habitats \citep{Hanski.1999b, MacArthur.2001}. However, the
ability to disperse depends on a sufficient degree of landscape connectivity
\citep{Fahrig.2003, Clobert.2012}, making the identification and protection of
dispersal corridors that promote connectivity a task of fundamental importance
\citep{Nathan.2008, Doerr.2011, Rudnick.2012}. Identifying dispersal corridors
not only necessitates a comprehensive understanding of the factors that limit
dispersal, but also an appropriate model to estimate connectivity
\citep{Baguette.2013, Vasudev.2015, Hofmann.2021}. To date, the most commonly
used connectivity models are least-cost path analysis (LCPA;
\citealp{Adriaensen.2003}) and circuit theory (CT; \citealp{McRae.2006,
McRae.2008}). Unfortunately, both models rest on assumptions that appear
unsuitable for dispersers, thus calling for the development of alternative
approaches. One promising alternative is to assess landscape connectivity via
simulated dispersal trajectories generated from individual-based movement models
(IBMMs, \citealp{Diniz.2019}). However, IBMMs require a large number of
subjective modeling decisions, thus rendering among-system comparisons
difficult.

@@

\subsection{Dispersal Simulation}
Dispersal simulations based on the most parsimonious movement model proved
useful for assessing landscape connectivity. Of the 50,000 simulated dispersal
trajectories with starting point within the main study area, only 4.5\% reached
a map boundary, suggesting that we successfully mitigated biases from boundary
effects. Moreover, our examination of the relative traversal frequency across
all checkpoints showed that the relative traversal frequency reached a steady
state after 10,500 simulated dispersal trajectories (Figure S3). Although
variability in the relative traversal frequency kept decreasing as we increased
the number of simulated dispersers, the marginal benefit of simulating
additional trajectories diminished quickly (Figure S3).

>>>>

\newpage
\begin{abstract}
%------------------------------------------------------------------------------
%  LONG VERSION
%------------------------------------------------------------------------------
\noindent 1. Dispersal of individuals contributes to long-term population
persistence, yet requires a sufficient degree of landscape connectivity. To
date, connectivity has mainly been investigated using least-cost analysis and
circuit theory, two methods that make assumptions that are hardly applicable to
dispersal. Least-cost analysis assumes that animals move towards a known
endpoint and are knowledgeable about the most favorable route. Circuit theory
presumes a complete random walk and fails to incorporate directional
persistence. While these assumptions can be relaxed by explicitly simulating
dispersal trajectories across the landscape, a unified approach for such
simulations is lacking.

\noindent 2. Here, we present a simple three-step approach to simulate dispersal
and to assess connectivity using empirical GPS movement data and a set of
habitat covariates. In step one, we use integrated step-selection functions to
fit a mechanistic movement model describing habitat and movement preferences of
dispersing individuals. In step two, we apply the parameterized model to
simulate dispersal across the study area. In step three, we derive a set of
three complementary connectivity maps; a heatmap highlighting frequently
traversed areas, a betweenness map pinpointing dispersal corridors, and a map of
inter-patch connectivity indicating the presence and intensity of functional
links between habitat patches. We demonstrate the applicability of the proposed
three-step approach in a case study in which we use GPS data collected on
dispersing African wild dogs (\textit{Lycaon pictus}) inhabiting the
Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA) in southern Africa.

\noindent 3. Using step-selection functions we successfully parametrized a
detailed dispersal model that described dispersing individuals habitat and
movement preferences, as well as potential interactions among the two. The model
substantially outperformed a model that omitted such interactions and enabled us
to simulate 80,000 dispersal trajectories across the study area.

\noindent 4. By explicitly simulating individuals, our approach not only
requires fewer unrealistic assumptions about dispersal, but also permits the
calculation of a variety of connectivity metrics that together provide a
comprehensive view of landscape connectivity. In our case study, the three
derived connectivity maps revealed several wild dog dispersal hotspots and
corridors across the extent of our study area, yet each map highlighted a
different aspect of landscape connectivity, thus emphasizing their complementary
nature. Overall, our case study demonstrates that a simulation-based approach
offers a simple yet powerful alternative to traditional connectivity modeling
techniques. Our approach is therefore useful for a variety of applications in
ecological, evolutionary, and conservation research.

@@

% Importance of Dispersal & Connectivity
\subsection{Importance of Connectivity \& Connectivity Models}
Dispersal of individuals is a vital process that allows species to maintain
genetic diversity \citep{Perrin.1999, Perrin.2000, Frankham.2002, Leigh.2012,
Baguette.2013}, rescue non-viable populations \citep{Brown.1977}, and to
colonize unoccupied habitats \citep{Hanski.1999b, MacArthur.2001}. However, the
ability to disperse depends on a sufficient degree of landscape connectivity
\citep{Fahrig.2003, Clobert.2012}, making the identification and protection of
dispersal corridors that promote connectivity a task of fundamental importance
\citep{Nathan.2008, Doerr.2011, Rudnick.2012}. Identifying dispersal corridors
not only necessitates a comprehensive understanding of the factors that limit
dispersal, but also an appropriate model to estimate connectivity
\citep{Baguette.2013, Vasudev.2015, Hofmann.2021}. To date, the most commonly
used connectivity models are least-cost path analysis (LCPA;
\citealp{Adriaensen.2003}) and circuit theory (CT; \citealp{McRae.2006,
McRae.2008}). Unfortunately, both models rest on assumptions that appear
unsuitable for dispersers, thus calling for the development of alternative
approaches. One promising alternative is to assess landscape connectivity via
simulated dispersal trajectories generated from individual-based movement models
(IBMMs, \citealp{Diniz.2019}). However, IBMMs require a large number of
subjective modeling decisions, thus making among-system comparisons difficult.

@@

\subsection{Dispersal Simulation}
Dispersal simulations based on the most parsimonious movement model proved
useful for assessing landscape connectivity. Of the 50,000 simulated dispersal
trajectories that originated from the main study area, only 4.5\% reached a map
boundary, suggesting that we successfully mitigated biases from boundary
effects. Moreover, our examination of the relative traversal frequency across
all checkpoints showed that the relative traversal frequency reached a steady
state after 10,500 simulated dispersal trajectories (Figure S3). Although
variability in the relative traversal frequency kept decreasing as we increased
the number of simulated dispersers, the marginal benefit of simulating
additional trajectories diminished quickly (Figure S3).

### 2022-02-22 Chapter_4/04_Manuscript/Manuscript.tex
updates

Step selection analysis is a powerful framework that enables researchers to
study habitat and movement preferences of their focal species. The method works
by comparing spatial covariates at locations where the studied animal has been
observed, to a set of locations where the animal was presumably absent. For
this, the collected GPS data is converted into steps, where a step represents
the straight line movement between two consecutive GPS fixes. One can then
compare covariates experienced along observed steps to covariates along
potential alternative steps and thus infer preferred or avoided features. By
restricting the availability domain to a set of steps, SSA is specifically
intended to account for spatio-temporal autocorrelation inherent to data
collected using GPS collars and is therefore one of the preferred methods for
analysing such data. However, it is considered good practice to only consider
steps with similar step durations in step-selection models. Depending on the
amount of available data and depending on the number of missing GPS fixes, this
might entail little sacrifice. In some situations, however, removing irregular
data implies a substantial loss of information. Additionally, recent
developments in the SSA realm have brought forward several improvements to SSA
that might allow to relax the assumption of regular step-durations. This
includes the \textit{integrated} SSA approach presented by \citep{Avgar.2016},
in which inference on habitat \textit{and} movement preferences are possible
thanks to the inclusion of movement descriptors in the respective model. More
recently, the method has been further refined by \citep{Munden.2021}, who coined
the term \textit{time-varying integrated} SSA. This method was developed for
high frequency data which has been rarified using a change-point detection
algorithm, thus also resulting in irregular step-durations.

@@

\subsubsection{Movement Simulation}
To simulate movement across the above generated virtual lanscape, we employed an
``inverted'' step-selection function, which proceeded as follows. First, we
generated a random starting location. To prevent starting points in the vicinity
of map borders, we restricted sampled locations to x- and y-coordinates between
50 and 250 (white dotted rectangle in \Cref{Covariates}). Second, we generated a
set of 10 random steps originating from the sampled starting point. We then
generated random steps by sampling turing angles from a von Mises distribution
with concentration parameter \(\kappa = 0.5\) and location parameter \( \mu = 0
\), and step lengths from a gamma distribution with shape parameter \(k = 3 \)
and scale parameter \(\theta = 1\). Third, we extracted covariate values along
each random step from the underlying covariate layers and compute the average of
each covariate along every step. Fourth, we predicted the probability of being
chosen for each step by applying \Cref{EQ1}. The vector of relative habitat
preferences used to make predictions was set to \(\beta_{water} = -1\),
\(\beta_{elev} = 0.5\), and \(\beta_{dist} = -15\). Fifth, we sampled one of the
random steps based on predicted probabilities and computed the new position of
the simulated individual. We the repated these steps until a total of 100 steps
were realized and we replicated the simulation 100 times, thus resulting in 100
independent movement trajectories (\Cref{Simulations}). Note that by excluding
interactions among habitat covariates and step metrics we implicitly assume
independent habitat and movement kernels \citep{Avgar.2016}.

\begin{figure}
  \begin{center}
  \includegraphics[width = 0.5\textwidth]{99_Simulations.png}
  \caption{100 Simulated movement trajectories, each comprising of 100 steps.
  Simulated individual were initiated within the blue rectangle to mitigate edge
  effects.}
  \label{Simulations}
  \end{center}
\end{figure}

@@

\paragraph{Data Rarification}
We rarified the simulated GPS data by randomly removing a fixed fraction of GPS
fixes. To assess the impact of different degrees of  ``missingness'', we
incrementally increased the fraction of missing fixes from 0.0 (complete
dataset) to 0.8 (80\% missing) with 10\% increments. The removal of GPS fixes
introduced irregular temporal intervals between remaining fixes; we will refer
to these intervals as ``step-times'', where the  ``step'' refers to the straight
line movement between two consecutive GPS fixes \citep{Turchin.1998}. In the
complete dataset, a step-time of one was assumed, whereas the step-time
increased by one for every fix that was missing between subsequent fixes. For
instance, a single missing fix between two other fixes directly increased the
step time from one to two.

@@

\paragraph{Step Length and Turning Angle Distributions}
To generate random steps, we employed two competing approaches. In the first
approach, we sampled step-lengths and turning-angles from the same distributions
that were used to simulate GPS data (cfr. Section ...). To adjust sampled step
lengths to the different step durations of observed steps, we multiplied sampled
step lengths with the step-duration of the respective observed step. However, in
a pre-analysis we found that step-lengths do not linearly scale with step
duration, hence we also tested an alternative approach. Specifically, we fitted
separate step-length and turning angle distributions for each step duration.

>>>>

Step selection analysis is a powerful framework that enables researchers to
study habitat and movement preferences of their focal species. The method works
by comparing spatial covariates at locations where the studied animal has been
observed, to a set of locations where the animal was presumably absent. For
this, the collected GPS data is converted into steps, where a step represents
the straight line movement between two consecutive GPS fixes. One can then
compare covariates experienced along observed steps to covariates along
potential alternative steps and thus infer preferred or avoided features. By
restricting the availability domain to a set of steps, SSA is specifically
intended to account for spatio-temporal autocorrelation inherent to data
collected using GPS collars and is therefore one of the preferred methods for
analysing such data. However, it is considered good practice to only consider
steps with similar step durations in step-selection models. Depending on the
amount of available data and depending on the number of missing GPS fixes, this
might entail little sacrifice. In some situations, however, removing irregular
data implies a substantial loss of information. Additionally, recent
developments in the SSA realm have brought forward several improvements to SSA
that might allow to relax the assumption of regular step-durations. This
includes the \textit{integrated} SSA approach presented by \citep{Avgar.2016},
in which inference on habitat \textit{and} movement preferences are possible
thanks to the inclusion of movement descriptors in the respective model. More
recently, the method has been further refined by \citep{Munden.2021}, who coined
the term \textit{time-varying integrated} SSA. This method was developed for
high frequency data which has been rarified using a change-point detection
algorithm, thus also resulting in irregular step-durations. In case of missing
steps, in contrast, steps durations are not entirely random but usually
clustered.

@@

\subsubsection{Movement Simulation}
To simulate movement across the above generated virtual lanscape, we employed an
``inverted'' step-selection function, which proceeded as follows. First, we
generated a random starting location. To prevent starting points in the vicinity
of map borders, we restricted sampled locations to x- and y-coordinates between
50 and 250 (white dotted rectangle in \Cref{Covariates}). Second, we generated a
set of 10 random steps originating from the sampled starting point. We then
generated random steps by sampling turning angles from a von Mises distribution
with concentration parameter \(\kappa = 0.5\) and location parameter \( \mu = 0
\), and step lengths from a gamma distribution with shape parameter \(k = 3 \)
and scale parameter \(\theta = 1\). Third, we extracted covariate values along
each random step from the underlying covariate layers and computed the average
of each covariate along every step. Fourth, we predicted for each step the
probability of being chosen by applying \Cref{EQ1}. The vector of relative
habitat preferences used to make predictions was set to \(\beta_{water} = -1\),
\(\beta_{elev} = 0.5\), and \(\beta_{dist} = -15\). Fifth, we sampled one of the
random steps based on predicted probabilities and computed the new position of
the simulated individual. We the repated these steps until a total of 100 steps
were realized and we replicated the simulation 100 times, thus resulting in 100
independent movement trajectories (\Cref{Simulations}). Note that by excluding
interactions among habitat covariates and step metrics we implicitly assume
independent habitat and movement kernels \citep{Avgar.2016}.

\begin{figure}
  \begin{center}
  \includegraphics[width = 0.5\textwidth]{99_Simulations.png}
  \caption{100 Simulated movement trajectories, each comprising of 100 steps.
  Simulated individuals were initiated within the blue rectangle to mitigate
  edge effects.}
  \label{Simulations}
  \end{center}
\end{figure}

@@

\paragraph{Data Rarefication}
We rarefied the simulated GPS data by randomly removing a fixed fraction of GPS
fixes. To assess the impact of different degrees of  ``missingness'', we
incrementally increased the fraction of missing fixes from 0.0 (complete
dataset) to 0.8 (80\% missing) with 10\% increments. The removal of GPS fixes
introduced irregular temporal intervals between remaining fixes; we will refer
to these intervals as ``step-times'', where the  ``step'' refers to the straight
line movement between two consecutive GPS fixes \citep{Turchin.1998}. In the
complete dataset, a step-time of one was assumed, whereas the step-time
increased by one for every fix that was missing between subsequent fixes. For
instance, a single missing fix between two other fixes directly increased the
step time from one to two.

@@

\paragraph{Step Length and Turning Angle Distributions}
To generate random steps, we employed three competing approaches, coined
\textit{uncorrected}, \textit{na\"ive}, and \textit{dynamic} approach,
respectively. In the uncorrected approach, we sampled step-lengths and
turning-angles from parametric gamma and von Mises distributions that were
fitted to observed steps with step-duration of one. In the \textit{na\"ive}
approach, we sampled step lengths and turning angles from the same
distributions, yet we linearly scaled sampled step lengths to the actual
step-duration. For instance, for any step with a step-duration of two, we
doubled the sampled step length. This approach is representative of the
time-varying SSF approach proposed by \cite{Munden.2021}. Finally, in the
\textit{dynamic} approach, we sampled step lengths and turning angles from
distributions that were fit to different step-durations. That is, we fitted
separate step length and turning angles distributions to every possible step
duration. To achieve this, we subsampled the observed data to 50\% of the fixes,
which resulted in varying step-durations. We then and fitted step length and
turning angle distributions to the steps from different step-durations. We
repeated this procedure 1000 times and averaged the parameter estimates across
replicates.
